---
title: "E-Rate Trends"
author: "Nathan Lovin"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    social: menu
    source_code: embed
---


```{r setup, include=FALSE}
source("setup.R")


## ---------------------------
## check.packages function: install and load multiple R packages.
## Check to see if packages are installed. Install them if they are not, then load them into the R session.

check.packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

## Extra packages for the dashboard
packages<-c("extrafont", "plotly", "leaflet","flexdashboard")
check.packages(packages)

rm(check.packages, packages)





## -----------------------------
## Setup map data
# Call world shapefile from the rnaturalearth package as a sf object
states_shp <- ne_states(returnclass = "sf",
                       country = "United States of America")

# Setup projections
US.proj <- "+proj=aea +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m"

yoy.state_geo <- left_join(states_shp,yoy.state, by = c("postal" = "state"))

US_geo <- yoy.state_geo %>% 
  filter(name.y != "Alaska" & name.y != "Hawaii")


```

Declining Requests
=======================================================================

Row
-----------------------------------------------------------------------
### Entity type by Year, dropping Voice requests

```{r yoy.requests, echo=F}

### Requests by Year

p1<-ggplot(yoy,
       aes(x = funding_year,
           y = requests)) +
  geom_bar(fill = "darkseagreen4",
           stat = 'identity',
           show.legend = FALSE) +
  geom_text(aes(label = ifelse(is.na(yoy.req), "", comma(yoy.req))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  scale_y_continuous(label=comma) +
  labs(title = "Annual E-Rate Requests", 
       x = "", y = "Requests (N)",
       subtitle = "2016 - 2018",
       caption = "Values above bars represent Year-over-Year changes") +
  ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=13),
        plot.caption = element_text(size=7))

ggplotly(p1)

```

### Total Amount Requested for E-Rate Funding ($)
```{r yoy.requests2, echo=F}

### Requests by Year
p2<-ggplot(yoy,
       aes(x = funding_year,
           y = ammount.req)) +
  geom_bar(fill = "dodgerblue4",
           stat = 'identity',
           show.legend = FALSE) +
  geom_text(aes(label = ifelse(is.na(yoy.ammount), "", comma(yoy.ammount))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  scale_y_continuous(label=dollar_format()) +
  labs(title = "Total Amount Requested for E-Rate Funding ($)", 
       x = "", y = "ammount requested ($)",
       subtitle = "2016 - 2018",
       caption = "Values above bars represent Year-over-Year changes") +
  ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=13),
        plot.caption = element_text(size=7))


ggplotly(p2)

```

Row
-----------------------------------------------------------------------

### Requests by Year, excluding Voice and Connections
```{r yoy.requests.novoice.noconnection, echo=F}


### Requests by Year, dropping voice and connections
p2<-ggplot(yoy.no.voice.connections,
           aes(x = funding_year,
               y = requests)) +
  geom_bar(fill = "darkseagreen4",
           stat = 'identity',
           show.legend = FALSE) +
  geom_text(aes(label = ifelse(is.na(yoy.req), "", comma(yoy.req))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  scale_y_continuous(label=comma) +
  labs(title = "Annual E-Rate Requests, excluding Voice and Internal Connections", 
       x = "", y = "Requests (N)",
       subtitle = "Voice and Internal Connection requests excluded | 2016 - 2018",
       caption = "Values above bars represent Year-over-Year changes") +
  ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=13),
        plot.subtitle = element_text(size=12),
        plot.caption = element_text(size=7))
ggplotly(p2)

```

### Amount by Year, dropping voice
```{r yoy.requests.novoice.noconnection2, echo=F}

### Amount by Year, dropping voice
p2<-ggplot(yoy.no.voice.connections,
       aes(x = funding_year,
           y = ammount.req)) +
  geom_bar(fill = "dodgerblue4",
           stat = 'identity',
           show.legend = FALSE) +
  geom_text(aes(label = ifelse(is.na(yoy.ammount), "", comma(yoy.ammount))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  scale_y_continuous(label=dollar_format()) +
  labs(title = "Total Amount Requested for E-Rate Funding ($), excluding Voice and Internal Connections", 
       x = "", y = "ammount requested ($)",
       subtitle = "Voice and Internal Connection requests excluded | 2016 - 2018",
       caption = "Values above bars represent Year-over-Year changes") +
  ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=13),
        plot.subtitle = element_text(size=11),
        plot.caption = element_text(size=7))

ggplotly(p2)

```


Requests by Service
=======================================================================

Row
-----------------------------------------------------------------------

### E-rate Requests by Service Type
```{r service.yoy, echo=F}

### Service Type Requests by Year
p1<-ggplot(service,
       aes(x = factor(form_471_service_type_name),
           y = requests,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.requests))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.requests), "", percent(yoy_pct.requests))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 2.5) +
  scale_fill_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title = "Annual Requests by Service Type", 
       x = "Service Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=comma) +
  ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=16),
        plot.caption = element_text(size=7))

ggplotly(p1)

```


Row
-----------------------------------------------------------------------

### Total Request Amounts by Service Type

```{r service.yoy2, echo=F}

### Total Request Amounts by Service Type
p1<-ggplot(service,
       aes(x = factor(form_471_service_type_name),
           y = dollars,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.dollars))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.dollars), "", percent(yoy_pct.dollars))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 2.5) +
  scale_fill_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title = "Total Amount Requested by Service", 
       x = "Service Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=dollar_format()) +
  ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=16),
        plot.caption = element_text(size=7))

ggplotly(p1)

```

Requests by Org. Type
=======================================================================

Row
-----------------------------------------------------------------------

### E-rate Requests by Organization Type

```{r entity.yoy, echo=F}
### Entity type by Year
p<-ggplot(entity,
       aes(x = factor(organization_entity_type_name),
           y = requests,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.requests))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.requests), "", percent(yoy_pct.requests))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 2.5) +
  scale_fill_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Requests by Organization Type", 
       x = "Organization Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=comma) +
    ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=16),
        plot.caption = element_text(size=7))
ggplotly(p)
```


Row
-----------------------------------------------------------------------

### Mean dollars requested by entity type and year

```{r}
p<-ggplot(entity,
       aes(x = factor(organization_entity_type_name),
           y = dollars,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.dollars))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.dollars), "", percent(yoy_pct.dollars))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 2.5) +
  scale_fill_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Total Amount Requested by Organization Type ($)", 
       x = "Organization Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=dollar_format()) +
    ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=16),
        plot.caption = element_text(size=7))
ggplotly(p)
```



Requests excl. Voice & Connections
=======================================================================

Row
-----------------------------------------------------------------------

### E-rate Requests by Service Type, Excluding Voice and Connections
```{r e.rate_no_n, echo=F}

### Entity type by Year, dropping voice and connections
p<-ggplot(entity.no_voice.connections,
       aes(x = factor(organization_entity_type_name),
           y = requests,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.requests))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.requests), "", percent(yoy_pct.requests))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 2.5) +
  scale_fill_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Requests by Organization Type, Voice and Internal Connection requests excluded", 
       x = "Organization Type", fill = "Funding Year", 
       subtitle = "Voice and Internal Connection requests excluded | 2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=comma) +
    ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=16),
        plot.subtitle = element_text(size=12),
        plot.caption = element_text(size=7))

ggplotly(p)

```


Row
-----------------------------------------------------------------------

### Total Request Amounts by Service, Excluding Voice and Connections

```{r e.rate_no_dollar, echo=F}

p1<-ggplot(entity.no_voice.connections,
       aes(x = factor(organization_entity_type_name),
           y = dollars,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.dollars))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.dollars), "", percent(yoy_pct.dollars))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 2.5) +
  scale_fill_ipsum(labels = function(x) str_wrap(x, width = 25)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Total Request Amount by Service ($), Voice and Internal Connection requests excluded", 
       x = "Service Type", fill = "Funding Year", 
       subtitle = "Voice and Internal Connection requests excluded | 2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label= dollar_format()) + 
  ggthemes::theme_fivethirtyeight() +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=16),
        plot.subtitle = element_text(size=12),
        plot.caption = element_text(size=7))

ggplotly(p1)

```


Requests by Service and Org.
=======================================================================

Row
-----------------------------------------------------------------------

### Requests by Service and Org.
```{r e.rate_org_serv.n, echo=F}

## REQUESTS
p<-ggplot(data = entity.service) +
  geom_bar(aes(x = factor(funding_year),
               y = requests,
               fill = factor(form_471_service_type_name)),
           position = "dodge",
           stat = 'identity') +
  scale_colour_discrete(labels = function(x) str_wrap(x, width = 5)) +
  ggthemes::theme_fivethirtyeight() + 
  scale_fill_ipsum(labels = function(x) str_wrap(x, width = 25)) + 
  facet_wrap(organization_entity_type_name~.,
             scales = "free") +
  labs(title = "Annual Requests by Service and Organization", 
       x = "", fill = "Service Requested", 
       subtitle = "2016-2018") +
  guides(fill=guide_legend(
    keywidth=0.2,
    keyheight=0.4,
    default.unit="inch")
  ) + 
  scale_y_continuous(label=comma) +
  theme(legend.text=element_text(size=7),
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        legend.title = element_text(size=8),
        plot.title = element_text(size=16),
        plot.subtitle = element_text(size=12),
        plot.caption = element_text(size=7))



ggplotly(p)

```

Amounts by Service and Org.
=======================================================================

Row
-----------------------------------------------------------------------

### Total Amounts Requested by Service and Org.

```{r e.rate_org_serv.dollar, echo=F}
## DOLLARS
p1<- entity.service %>% 
  mutate(dollars = dollars / 10^6) %>% 
ggplot() +
  geom_bar(aes(x = factor(funding_year),
               y = dollars,
               fill = factor(form_471_service_type_name)),
           position = "dodge",
           stat = 'identity', show.legend = F) +
  scale_colour_discrete(labels = function(x) str_wrap(x, width = 5)) +
  ggthemes::theme_fivethirtyeight() + 
  scale_fill_ipsum(labels = function(x) str_wrap(x, width = 25)) + 
  facet_wrap(organization_entity_type_name~.,scales = "free") +
  labs(title = "Amount Requested by Service and Entity ($)", 
       x = "", fill = "Service Requested", 
       subtitle = "2016-2018") +
  scale_y_continuous(label=dollar_format(suffix = "M"))

ggplotly(p1)

```



Maps
=======================================================================

Column {.tabset}
-------------------------------------
    
### Map 1: E-Fund Dollars Requested per Person, 2016
    
```{r}
m <- US_geo %>% 
  filter(funding_year==2016) %>% 
  select(ammount.per.person,everything()) %>% 
  tm_shape(., projection = US.proj) +
  tm_layout(frame = FALSE,
            main.title = "E-Fund Dollars Requested per Person, 2016") +
  tm_fill(col = "ammount.per.person",
          breaks = c(1,5,8,10,20,320),
          title = "Dollars Requested per Person") +
  tm_borders()
tmap_leaflet(m)
```

### Map 2: E-Fund Dollars Requested per Person, 2017

```{r}
m <- US_geo %>% 
  filter(funding_year==2017) %>% 
  select(ammount.per.person,everything()) %>% 
  tm_shape(., projection = US.proj) +
  tm_layout(frame = FALSE,
            main.title = "E-Fund Dollars Requested per Person, 2017") +
  tm_fill(col = "ammount.per.person",
          breaks = c(1,5,8,10,20,320),
          title = "Dollars Requested per Person") +
  tm_borders()
tmap_leaflet(m)
```

### Map 3: E-Fund Dollars Requested per Person, 2018

```{r}
m <- US_geo %>% 
  filter(funding_year==2018) %>% 
  select(ammount.per.person,everything()) %>% 
  tm_shape(., projection = US.proj) +
  tm_layout(frame = FALSE,
            main.title = "E-Fund Dollars Requested per Person, 2018") +
  tm_fill(col = "ammount.per.person",
          breaks = c(1,5,8,10,20,320),
          title = "Dollars Requested per Person") +
  tm_borders()
tmap_leaflet(m)

# breaks = c(1,5,8,10,20,320)

```