---
title: "test2"
author: "Nathan Lovin"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    social: menu
    source_code: embed
---

```{r setup, include=FALSE}
library(rlang)
library(scales)
library(viridisLite)
library(fs)
library(hrbrthemes)
library(RSocrata)
library(plotly)
library(noncensus)
library(sp)
library(sf)
library(rnaturalearth)
library(tmap)
library(leaflet)
library(grid)
library(tidyverse)
library(flexdashboard)


load("Data/FRN_status_raw.Rdata")

frn <- FRN.status %>% 
  filter(between(funding_year,2016,2018))

## YoY Calculation
yoy <- frn %>% 
  group_by(funding_year) %>% 
  summarise(requests = n(),
            ammount.req = sum(funding_commitment_request,na.rm = T)) %>% 
  mutate(yoy.req = requests - lag(requests),
         yoy_pct.req = round(100*((requests - lag(requests))/requests),1),
         yoy.ammount = ammount.req - lag(ammount.req),
         yoy_pct.ammount = round(100*((ammount.req - lag(ammount.req))/ammount.req),1))

## Load state pop data
data(states)  

yoy.state <- frn %>% 
  group_by(state,funding_year) %>% 
  summarise(requests = n(),
            ammount.req = sum(funding_commitment_request, na.rm = T)) %>% 
  mutate(yoy.req = requests - lag(requests),
         yoy_pct.req = round(100*((requests - lag(requests))/requests),1),
         yoy.ammount = ammount.req - lag(ammount.req),
         yoy_pct.ammount = round(100*((ammount.req - lag(ammount.req))/ammount.req),1)) %>% 
  ungroup() %>% 
  left_join(., states, by = "state") %>% 
  group_by(funding_year) %>% 
  mutate(total.requests_year = sum(requests),
         total.ammount_year = sum(ammount.req),
         population = as.numeric(population)) %>% 
  ungroup() %>% 
  group_by(state, funding_year) %>% 
  mutate(pct_of_total.req = 100*(requests/total.requests_year),
         pct_of_total.ammount = 100*(ammount.req/total.ammount_year),
         req.state.pct = 100*(requests/population),
         ammount.per.person = round(ammount.req/population,0))

## Remove state pop data after merging
rm(states)

## ---------------------------
## Total Requests & Request Ammounts by Service Type
service <- frn %>% 
  group_by(form_471_service_type_name,funding_year) %>% 
  summarise(requests = n(),
            dollars = sum(funding_commitment_request, na.rm = T)) %>% 
  mutate(yoy.dollars = dollars - lag(dollars),
         yoy_pct.dollars = ((dollars - lag(dollars))/dollars),
         yoy.requests = requests - lag(requests),
         yoy_pct.requests = ((requests - lag(requests))/requests))

## Total Requests & Request Ammounts by Entity
entity <- frn %>% 
  group_by(organization_entity_type_name,funding_year) %>% 
  summarise(requests = n(),
            dollars = sum(funding_commitment_request, na.rm = T)) %>% 
  mutate(yoy.dollars = dollars - lag(dollars),
         yoy_pct.dollars = ((dollars - lag(dollars))/dollars),
         yoy.requests = requests - lag(requests),
         yoy_pct.requests = ((requests - lag(requests))/requests))

## Total Requests & Request Ammounts by Entity, dropping Voice requests
entity.no_voice <- frn %>% 
  filter(form_471_service_type_name != "Voice") %>% 
  group_by(organization_entity_type_name,funding_year) %>% 
  summarise(requests = n(),
            dollars = sum(funding_commitment_request, na.rm = T)) %>% 
  mutate(yoy.dollars = dollars - lag(dollars),
         yoy_pct.dollars = ((dollars - lag(dollars))/dollars),
         yoy.requests = requests - lag(requests),
         yoy_pct.requests = ((requests - lag(requests))/requests))



### ------------------------------------------------------------
## Maps

# Call world shapefile from the rnaturalearth package as a sf object
states_shp <- ne_states(returnclass = "sf",
                       country = "United States of America")


# Setup projections
US.proj <- "+proj=aea +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m"
AK.proj <- "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
HI.proj <- "+proj=aea +lat_1=8 +lat_2=18 +lat_0=13 +lon_0=-153 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"


yoy.state_geo <- left_join(states_shp,yoy.state, by = c("postal" = "state"))

US_geo <- yoy.state_geo %>% 
  filter(name.y != "Alaska" & name.y != "Hawaii")

AK_geo <- yoy.state_geo %>% 
  filter(name.y == "Alaska")

HI_geo <- yoy.state_geo %>% 
  filter(name.y == "Hawaii") %>% 
  st_crop(., c(xmin=-162.62, xmax=-153.87, ymin=18.05, ymax=23.68))


```

E-fund Requests (N)
=======================================================================

Row
-----------------------------------------------------------------------

### Service Type Requests by Year

```{r}
p <- ggplot(service,
       aes(x = factor(form_471_service_type_name),
           y = requests,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.requests))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.requests), "", percent(yoy_pct.requests))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -1,    # nudge above top of bar
            size = 3) +
  scale_fill_ipsum() +
  theme_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Requests by Service Type", 
       x = "Service Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=comma)

ggplotly(p)
```


### Entity type by Year

```{r}
p <- ggplot(entity,
       aes(x = factor(organization_entity_type_name),
           y = requests,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.requests))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.requests), "", percent(yoy_pct.requests))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  scale_fill_ipsum() +
  theme_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Requests by Organization Type", 
       x = "Organization Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=comma)

ggplotly(p)
```

Row
-----------------------------------------------------------------------

### Entity type by Year, dropping Voice requests

```{r}
p <- ggplot(entity.no_voice,
       aes(x = factor(organization_entity_type_name),
           y = requests,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.requests))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.requests), "", percent(yoy_pct.requests))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  scale_fill_ipsum() +
  theme_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Requests by Organization Type", 
       x = "Organization Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=comma)            # Add a loess smoothed fit curve with confidence region
ggplotly(p)
```

### Constraining Slope with stat_smooth

```{r}
p <- ggplot(data = frn) +
  geom_bar(aes(x = organization_entity_type_name, fill = factor(form_471_service_type_name)),position = "dodge") +
  theme_ipsum() + 
  scale_fill_ipsum(labels = function(x) str_wrap(x, width = 25)) + 
  facet_grid(funding_year~.) +
  labs(title = "Requests by Organization Type", 
       x = "Organization Type", fill = "Service Requested", 
       subtitle = "2016-2018") +
  guides(fill=guide_legend(
    keywidth=0.2,
    keyheight=0.4,
    default.unit="inch")
  )
ggplotly(p)
```

Amounts Requested ($)
=======================================================================

Row
-----------------------------------------------------------------------

### Total Request Ammounts by Service Type

```{r}
p <- ggplot(service,
       aes(x = factor(form_471_service_type_name),
           y = dollars,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.dollars))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.dollars), "", percent(yoy_pct.dollars))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  scale_fill_ipsum() +
  theme_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Total Request Amount by Service", 
       x = "Service Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label=dollar_format())
ggplotly(p)
```

### Total Request Amounts by Entity

```{r}
p <- ggplot(entity,
       aes(x = factor(organization_entity_type_name),
           y = dollars,
           fill = factor(funding_year),
           group = factor(funding_year),
           label = percent(yoy_pct.dollars))) +
  geom_bar(position = "dodge",
           stat = "identity") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.dollars), "", percent(yoy_pct.dollars))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  scale_fill_ipsum() +
  theme_ipsum() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  labs(title = "Total Request Amount by Service", 
       x = "Service Type", fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  scale_y_continuous(label= dollar_format())

ggplotly(p)
```

Row
-----------------------------------------------------------------------

### Mean dollars requested by entity type and year

```{r}
p <- frn %>% 
  group_by(organization_entity_type_name, funding_year) %>% 
  summarise(mean_request = mean(funding_commitment_request, na.rm = T)) %>% 
  mutate(yoy.ammount = mean_request - lag(mean_request),
         yoy_pct.ammount = ((mean_request - lag(mean_request))/mean_request)) %>% 
  ggplot(aes(x = organization_entity_type_name, ## --- ggplot begins here --- ##
             y = mean_request, 
             fill = factor(funding_year), 
             group = factor(funding_year),
             label = percent(yoy_pct.ammount))) + 
  geom_bar(stat = "identity", 
           position = "dodge") +
  labs(title = "Average Request Amount ($) by Organization", 
       x = "Organization Type", 
       y = "Avg. Request", 
       fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.ammount), "", percent(yoy_pct.ammount))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  theme_ipsum() +
  scale_fill_ipsum() +
  scale_y_continuous(labels = dollar_format())
ggplotly(p)
```

### Mean dollars requested by service type and year

```{r}
p <- frn %>% 
  group_by(form_471_service_type_name, funding_year) %>% 
  summarise(mean_request = mean(funding_commitment_request, na.rm = T)) %>% 
  mutate(yoy.ammount = mean_request - lag(mean_request),
         yoy_pct.ammount = ((mean_request - lag(mean_request))/mean_request)) %>% 
  ggplot(aes(x = form_471_service_type_name, ## --- ggplot begins here --- ##
             y = mean_request, fill = factor(funding_year), 
             group = factor(funding_year),
             label = percent(yoy_pct.ammount))) + 
  geom_bar(stat = "identity", 
           position = "dodge") +
  labs(title = "Average Request Amount ($) by Service", 
       x = "Service Type", 
       y = "Avg. Request", 
       fill = "Funding Year", 
       subtitle = "2016-2018",
       caption = "Percentages represent Year-over-Year changes") +
  geom_text(aes(label = ifelse(is.na(yoy_pct.ammount), "", percent(yoy_pct.ammount))),
            position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) +
  theme_ipsum() +
  scale_fill_ipsum() +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20))

ggplotly(p)
```

Geographic Inisghts
=======================================================================

Column {.tabset}
-------------------------------------
    
### Map 1: E-Fund Dollars Requested per Person, 2016
    
```{r}
m <- US_geo %>% 
  filter(funding_year==2016) %>% 
  select(ammount.per.person,everything()) %>% 
  tm_shape(., projection = US.proj) +
  tm_layout(frame = FALSE,
            main.title = "E-Fund Dollars Requested per Person, 2016") +
  tm_fill(col = "ammount.per.person",
          breaks = c(1,5,8,10,20,320),
          title = "Dollars Requested per Person") +
  tm_borders()
tmap_leaflet(m)
```

### Map 2: E-Fund Dollars Requested per Person, 2017

```{r}
m <- US_geo %>% 
  filter(funding_year==2017) %>% 
  select(ammount.per.person,everything()) %>% 
  tm_shape(., projection = US.proj) +
  tm_layout(frame = FALSE,
            main.title = "E-Fund Dollars Requested per Person, 2017") +
  tm_fill(col = "ammount.per.person",
          breaks = c(1,5,8,10,20,320),
          title = "Dollars Requested per Person") +
  tm_borders()
tmap_leaflet(m)
```

### Map 3: E-Fund Dollars Requested per Person, 2018

```{r}
m <- US_geo %>% 
  filter(funding_year==2018) %>% 
  select(ammount.per.person,everything()) %>% 
  tm_shape(., projection = US.proj) +
  tm_layout(frame = FALSE,
            main.title = "E-Fund Dollars Requested per Person, 2018") +
  tm_fill(col = "ammount.per.person",
          breaks = c(1,5,8,10,20,320),
          title = "Dollars Requested per Person") +
  tm_borders()
tmap_leaflet(m)

# breaks = c(1,5,8,10,20,320)

```